From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-02-27 appendixB/appendixB.Rmd
environmental noise

preamble: "\\newcommand{\\ud}{\\mathrm{d}}"
---

@@

\begin{align}
\frac{\ud p_n}{\ud t} = (\mathbb E^{-1} - 1) b_n p_n + (\mathbb E -1) d_n p_n \label{master1}
\end{align}

@@

\begin{align}
\frac{\ud }{\ud t} \langle n \rangle &= \sum n (\mathbb E^{-1} - 1) b_n p_n + \sum n (\mathbb E -1) d_n p_n \nonumber \\ 
&= \sum  b_n p_n(\mathbb E - 1)n  + \sum d_n p_n(\mathbb E^{-1} -1) n \nonumber \\
&= - \langle d_n \rangle +  \langle b_n \rangle \label{mean}
\end{align}

@@

\begin{align}
\frac{\ud }{\ud t} \langle n^2 \rangle &= 2 \langle n (b_n - d_n) \rangle + \langle b_n+d_n \rangle \label{var}
\end{align}

@@

\begin{equation}
\ud X_t = \left[ b(x) - d(x) \right] \ud t + \sqrt{b(X_t) + d(X_t)} \ud W_t
\end{equation}

@@

\begin{align}
\frac{\ud n}{\ud t} &= \Omega \frac{\ud \phi(t)}{\ud t}  + \Omega^{1/2}\frac{\ud \xi}{\ud t} = 0, \nonumber \\
\intertext{therefore}
\frac{\ud \xi}{\ud t} &= -\Omega^{1/2}\frac{\ud \phi(t)}{\ud t}. \label{dxidt}
\end{align}

Also by the chain rule, we have
\begin{align*}
\frac{\partial}{\partial t} P(n,t) &= \frac{\partial \Pi}{\partial t}  + \frac{\partial \Pi}{\partial \xi}\frac{\ud \xi}{\ud t} 
\end{align*}
Substituting in \eqref{dxidt}, we find:

\begin{align}
\frac{\partial}{\partial t} P(n,t) &= \frac{\partial \Pi}{\partial t}  - \Omega^{1/2}\frac{\ud \phi}{\ud t}\frac{\partial \Pi}{\partial \xi} \label{Pi}
\end{align}

@@

\begin{align}
&\frac{\partial \Pi(\xi, t)}{\partial t}  - \Omega^{1/2}\frac{\ud \phi}{\ud t}\frac{\partial \Pi}{\partial \xi}  = \nonumber \\
&\quad \left( -\Omega^{-1/2} \frac{\partial}{\partial \xi} +  \frac{\Omega^{-1}}{2} \frac{\partial^2}{\partial \xi^2} + \ldots\right)b(\phi\Omega+\xi\Omega^{1/2})\Pi(\xi, t)\nonumber \\
+&\quad \left( \Omega^{-1/2} \frac{\partial}{\partial \xi} +  \frac{\Omega^{-1}}{2} \frac{\partial^2}{\partial \xi^2} + \ldots\right)d(\phi\Omega+\xi\Omega^{1/2})\Pi(\xi, t)
\end{align}

@@

\begin{align}
\frac{\ud \phi(t)}{\ud t} = b(\phi)-d(\phi) = a_1(\phi) \label{macroscopic}
\end{align}

@@

This expression has the wonderful properties of being both simple and general.

@@

\begin{align}
\frac{\ud n}{\ud t} = c n \left(1 - \frac{n}{N}\right) - e n \label{levins}
\end{align}

@@

$$\frac{\ud n}{\ud t} = r n \left( 1 - \frac{n}{K} \right)$$
though the partition into birth and death rates is not explicit.  Different ways of dividing this equation between birth and death can therefore create different fluctuation patterns, even as the macroscopic average remains unchanged.  For instance, taking $b = rn$ and $d = rn^2/K$ and  we can calculate the fluctuations of the logistic equation around its equilibrium $n^* = K$ as:

@@

We introduce environmental noise into the master equation by allowing the transition rates to depend explicitly on the macroscopic variable $\psi$ as well as the state variable $\phi$.  It can be helpful to think of fluctuations in $\psi$ as also arising from some lower level transitions, but we need not model these explicitly.  Following the expansion as before (see @vanKampen2007 for details), the mean dynamics are given by the macroscopic equations:

\begin{align*}
\frac{\ud \phi}{\ud t} &= a_1(\phi, \psi) \\
\frac{\ud \psi}{\ud t} &= \beta_1(\psi) 
\end{align*}

while the fluctuation dynamics are now given by three coupled differential equations:

\begin{align}
\frac{\ud \langle \xi^2 \rangle}{\ud t} &= 2 \frac{\partial a_1}{\partial \phi} \langle \xi^2 \rangle + 2 \frac{\partial a_1 }{\partial \psi} \langle \xi \eta \rangle + a_2 \label{phivar} \\
\frac{\ud \langle \xi \eta \rangle}{\ud t} &= \left( \frac{\partial a_1}{\partial \phi} + \frac{\partial \beta_1 }{\partial \psi} \right)\langle \xi \eta \rangle + \frac{\partial a_1 }{\partial \psi} \langle \eta^2 \rangle \label{covar}\\
\frac{\ud \langle \eta^2 \rangle}{\ud t} &= 2 \frac{\partial \beta_1}{\partial \psi} \langle \eta^2 \rangle + \beta_2 \label{psivar}
\end{align}

Without an explicit model for the environmental variation process, we can merely summarize its role in terms of the autocorrelation, $\tau_c := \tfrac{\partial \beta_1}{\partial \psi}$ and its overall variance at steady-state, $\langle \eta^2 \rangle = \tfrac{\beta_2}{-2 \partial_{\psi} \beta{1,0}}$, which we will refer to as the environmental variation $\sigma^2_e := \langle \eta^2 \rangle$.  A more detailed mechanistic description of how the environmental noise arises would permit a time-dependent solution, but the system size expansion tells us that these two summary statistics are all we need to express the steady-state solution above.

Setting the remaining equations to zero to solve for the stationary state we can write down a general formula for the variation in $\phi$:

\begin{equation*}
\langle xi^2 \rangle = \frac{ \left( \partial_{\psi} a_1(\phi,\psi) \right)^2}{\left( \partial_{\phi} a_1(\phi,\psi) \right)^2 + \partial_{\phi} a_1 (x,y) \tau_c } \sigma_e^2 + \sigma^2_d
\end{equation*}

<!-- FIXME -->

$$
\begin{align}
\frac{\mathrm{d} x}{\mathrm{d} t} &= \alpha_1(x, y)  \\
\frac{\mathrm{d} y}{\mathrm{d} t} &= \beta_1(y) 
\end{align}
$$

while the fluctuation dynamics are now given by three coupled differential equations:

For a generic birth death process that depends both on
on population state $x$ but also on environmental
factors $y$, where we write $a_1(x,y) := b(x,y) - d(x,y)$,
sum $a_2(x,y) := b(x,y) + d(x,y)$

For a generic population model
that depends not only on population state $x$ but also on environmental
factors $y$, it is possible to derive the steady-state variance
(Appendix B) as:

$$
\begin{align}
&\sigma_x^2  = \frac{  \partial_y a_1(x,y) ^2}{ \partial_x a_1(x,y) ^2 +  \tau_c \phantom\cdot \partial_x a_1(x,y)  } \sigma_e^2 + \frac{1}{2} \frac{a_2(x,y)}{\partial_x a_1(x,y)}
\end{align}
$$

where $\sigma_e^2$ is the variance in the environmental noise and
$\tau_c$ the auto-correlation of that noise.

>>>>

preamble: "\\newcommand{\ud}{\\mathrm{d}}"
---

@@

\begin{align}
\frac{\mathrm{d} p_n}{\mathrm{d} t} = (\mathbb E^{-1} - 1) b_n p_n + (\mathbb E -1) d_n p_n \label{master1}
\end{align}

@@

\begin{align}
\frac{\mathrm{d} }{\mathrm{d} t} \langle n \rangle &= \sum n (\mathbb E^{-1} - 1) b_n p_n + \sum n (\mathbb E -1) d_n p_n \nonumber \\ 
&= \sum  b_n p_n(\mathbb E - 1)n  + \sum d_n p_n(\mathbb E^{-1} -1) n \nonumber \\
&= - \langle d_n \rangle +  \langle b_n \rangle \label{mean}
\end{align}

@@

\begin{align}
\frac{\mathrm{d} }{\mathrm{d} t} \langle n^2 \rangle &= 2 \langle n (b_n - d_n) \rangle + \langle b_n+d_n \rangle \label{var}
\end{align}

@@

\begin{equation}
\mathrm{d} X_t = \left[ b(x) - d(x) \right] \mathrm{d} t + \sqrt{b(X_t) + d(X_t)} \mathrm{d} W_t
\end{equation}

@@

\begin{align}
\frac{\mathrm{d} n}{\mathrm{d} t} &= \Omega \frac{\mathrm{d} \phi(t)}{\mathrm{d} t}  + \Omega^{1/2}\frac{\mathrm{d} \xi}{\mathrm{d} t} = 0, \nonumber \\
\intertext{therefore}
\frac{\mathrm{d} \xi}{\mathrm{d} t} &= -\Omega^{1/2}\frac{\mathrm{d} \phi(t)}{\mathrm{d} t}. \label{dxidt}
\end{align}

Also by the chain rule, we have
\begin{align*}
\frac{\partial}{\partial t} P(n,t) &= \frac{\partial \Pi}{\partial t}  + \frac{\partial \Pi}{\partial \xi}\frac{\mathrm{d} \xi}{\mathrm{d} t} 
\end{align*}
Substituting in \eqref{dxidt}, we find:

\begin{align}
\frac{\partial}{\partial t} P(n,t) &= \frac{\partial \Pi}{\partial t}  - \Omega^{1/2}\frac{\mathrm{d} \phi}{\mathrm{d} t}\frac{\partial \Pi}{\partial \xi} \label{Pi}
\end{align}

@@

\begin{align}
&\frac{\partial \Pi(\xi, t)}{\partial t}  - \Omega^{1/2}\frac{\mathrm{d} \phi}{\mathrm{d} t}\frac{\partial \Pi}{\partial \xi}  = \nonumber \\
&\quad \left( -\Omega^{-1/2} \frac{\partial}{\partial \xi} +  \frac{\Omega^{-1}}{2} \frac{\partial^2}{\partial \xi^2} + \ldots\right)b(\phi\Omega+\xi\Omega^{1/2})\Pi(\xi, t)\nonumber \\
+&\quad \left( \Omega^{-1/2} \frac{\partial}{\partial \xi} +  \frac{\Omega^{-1}}{2} \frac{\partial^2}{\partial \xi^2} + \ldots\right)d(\phi\Omega+\xi\Omega^{1/2})\Pi(\xi, t)
\end{align}

@@

\begin{align}
\frac{\mathrm{d} \phi(t)}{\mathrm{d} t} = b(\phi)-d(\phi) = a_1(\phi) \label{macroscopic}
\end{align}

@@

This expression has the wonderful properties of being both simple and quite general.

@@

\begin{align}
\frac{\mathrm{d} n}{\mathrm{d} t} = c n \left(1 - \frac{n}{N}\right) - e n \label{levins}
\end{align}

@@

$$\frac{\mathrm{d} n}{\mathrm{d} t} = r n \left( 1 - \frac{n}{K} \right)$$
though the partition into birth and death rates is not explicit.  Different ways of dividing this equation between birth and death can therefore create different fluctuation patterns, even as the macroscopic average remains unchanged.  For instance, taking $b = rn$ and $d = rn^2/K$ and  we can calculate the fluctuations of the logistic equation around its equilibrium $n^* = K$ as:

@@

Environmental noise arises not as a summary of lower-level processes, like in the case of intrinsic noise, but rather as a recognition that things we treat as model parameters are in fact dynamic variables themselves and subject to their own dynamics which we summarize statistically rather than model explicitly.  We can recognize this as simply a special case of a multivariate system size expansion; e.g. where a parameter of our birth or death process varies according to some environmental dynamics (e.g. a death rate $e$ linked to fluctuations in temperature, $y$.)  In general, we can write this as a macroscopic equation in two variables:

$$\frac{\mathrm{d} x}{\mathrm{d} t} = a_1(x, y) $$

In general, we could imagine a specific equation for the dynamics of the environmental process $y$, perhaps also subject to a system size expansion. To keep things simple, we will adopt a linear OU process for the noise,

$$
\begin{align*}
\mathrm{d} y &= - \beta y \phantom \cdot \mathrm{d} t + \sigma_e \mathrm{d}W
\end{align*}
$$
which corresponds to mean-zero Gaussian noise with auto-correlation $\beta$, where the white-noise limit corresponds to $\beta \to 0$. Following the expansion as before (see @vanKampen2007 for details of the multi-variate case), the mean dynamics for the population size are once again given by the macroscopic equation, $\frac{\mathrm{d} x}{\mathrm{d} t} = a_1(x, y) $, while the fluctuation dynamics are now given by three coupled differential equations:

$$
\begin{align}
\frac{\mathrm{d} \langle \xi^2 \rangle}{\mathrm{d} t} &= 2 \frac{\partial a_1}{\partial x} \langle \xi^2 \rangle + 2 \frac{\partial a_1 }{\partial y} \langle \xi \eta \rangle + a_2 \\
\frac{\mathrm{d} \langle \xi \eta \rangle}{\mathrm{d} t} &= \left( \frac{\partial a_1}{\partial x} - \beta \right)\langle \xi \eta \rangle + \frac{\partial a_1 }{\partial y} \langle \eta^2 \rangle \\
\frac{\mathrm{d} \langle \eta^2 \rangle}{\mathrm{d} t} &= -2 \beta \langle \eta^2 \rangle + \gamma_e^2
\end{align}

$$

Note that the fluctuations in population density $x$ depends as before on the fluctuation-disspation tradeoff, $2 \frac{\partial a_1}{\partial x} \langle \xi^2 \rangle + a_2$, but include the additonal term arising from a covariance with the environmental flucatuations. We can simplify this expression as before by considering the equilibrium noise dynamics

$$
\langle \xi^2 \rangle = \underbrace{-\frac{a_2}{2\partial_x a_1}}_{\textrm{demographic noise}} + \underbrace{\frac{\left[\partial_{y} a_1\right] ^2}{\partial_x a_1 - \beta} \frac{\langle\eta^2\rangle}{2\partial_x a_1}}_{\textrm{environmental contribution}}
$$

Here we have the familiar demographic noise term, coupled to an environmental noise contribution which depends on the details of how the environmental noise enters the macroscopic model, $\partial_y a_1$, as well as the relative correlation times $\partial_x a_1$ and $\beta$ of the population process and environmental process.
Note that we have left this expression in terms of $\langle \eta^2 \rangle$, rather than subsituting in the equilibrium expression

$$\langle \eta^2 \rangle = \frac{\gamma_e^2}{2\beta},$$

so that we can consider separately the limit of white environmental noise without also increasing the variance of the process. Changing the correlation while holding the total variation $\langle \eta^2\rangle$ constant thus amounds to increasing the intrinsic noise level $\gamma^2$ proportionally.  In general, a more mechanistic description of what it means to change the autocorrelation of an environmental process for a given variance is needed.  Nevertheless, considering the the white noise limit with some fixed variance, when $\beta \ll \partial_x \alpha_1$,  this expression simplifies as:

$$
\langle \xi^2 \rangle = \underbrace{-\frac{a_2}{2\partial_x a_1}}_{\textrm{demographic noise}} +  \underbrace{\frac{1}{2} \left[\frac{\partial_{y} a_1}{ \partial_x a_1}\right] ^2}_{\textrm{environmental coupling}} \langle\eta^2\rangle
$$
Taking extinction as such a Gaussian white noise process with meant $\bar e$ and variance $\sigma_e^2$, we can plug in our birth-death equations for the Levins model (and rescaling by system size, from $x = n/N$ back to $n = xN$), into the above equation to find the expected fluctuations are:

$$\sigma_n^2 = \frac{\bar e}{c}N + \frac{\left[(1-\frac{\bar{e}}{c}) N\right]^2}{(\bar e-c)^2} \sigma_e^2$$

where we recognize the term $(1-\frac{\bar{e}}{c}) N$ as the equilibrium population size, $\bar n$.  Meanwhile, if timescale of environmental fluctuations is much longer than population fluctuations (a red noise limit, $\beta \gg \partial_x \alpha_1$) this can instead be written as

$$
\langle \xi^2 \rangle = \underbrace{-\frac{a_2}{2\partial_x a_1}}_{\textrm{demographic noise}} + \underbrace{- \frac{\left[\partial_{y} a_1\right] ^2}{2\partial_x a_1 \beta}}_{\textrm{environmental coupling}}  \langle\eta^2\rangle
$$
(Recall that we have taken these expressions at equilibrium, and at equilibrium the derivative of the macroscopic equation $\partial_x a_1 < 0$ by definition, which guarentees the variances above are strictly positive).  In this limit ($\beta \gg \partial_x \alpha_1$), the overall noise decreases both with stronger dampening / higher autocorelation in the environmental process, (larger $\beta$) just as it does with increased dampening in population dynamics $\partial_x a_1$.

<!--
## Crowley model

Consider the Crowley model \cite{Crowley05} of competition for space:

\begin{align}
\dot x &= b_1 x (1-x-y) - d_1 x + cxy \\
\dot y &= b_2 y (1-x-y) - d_2 y - cxy
\end{align}

Where $x$ and $y$ are the respective densities of two species competing for space.  In the model species $x$ is the better competitor, displacing species $y$ from patches at rate $cyx$, while species $y$ is the better colonizer when $b_2 > b_1$, taking $d_1 = d_2$ for simplicity.  Under this competition-colinization trade-off, this the deterministic skeleton supports stable coexistence of both species.  We intepret this as an individaul model analgous to the Levins model, with a finite number of patches $N$ that both species compete to occupy. Taking $N$ as our system size ($\Omega$), we have $n_x = \Omega x + \Omega^{1/2} \xi$ and $n_y = \Omega y + \Omega^{1/2} \eta$, we proceed much as before to derive the fluctuation equations:

$$
\begin{align*}
\partial_t \sigma_{\xi}^2 &= 2( b_1(1-2x-y) + c y-d_1 )\sigma_{\xi}^2 \\ 
& \quad + 2( -b_1x+cx )\langle x y \rangle		 \\
& \quad + b_1 x(1-x-y) + c x y+d_2 x)\\
\sigma_{\eta}^2 &= 2(b_2(1-2y-x) - c x-d_2)\sigma_{\eta}^2							\ \\
& \quad + 2(-b_2 y-c y)\langle \xi \eta \rangle								\\
& \quad + b_2 y(1-x-y) + c x y+d_2 y)\\
\langle \xi \eta \rangle &= b_1(1-2x-y)+c y-d_1)\langle \xi \eta \rangle			 \\
& \quad + (-b_1 x+c x) \sigma^2_{\eta}										 \\ 
& \quad + (-b_2 y - c y)\sigma^2_{\xi}										 \\ 
& \quad + (b_2(1-2 y-x) - c x - d_2)\langle \xi \eta \rangle - c x y;
\end{align*}
$$

These three coupled equations can be solved by numerical integration, which we compare to individual-based simulation of the Crowley competition model, Fig~\ref{crowley}.

\begin{figure}
\begin{center}
\includegraphics[width=.5\textwidth]{crowley_fluc.png}
\caption{Fluctuations for species $x$ in black and $y$ in red.  Smooth curves are theoretical predictions of variance within each population.  Model parameters: $b_1 = 0.2$, $b_2 = 0.6$, $d_1 = d_2 = c = 0.1$, $N=1000$, plots based on exact Gillespie simulations averaged over an ensemble of $10^2$ realizations.}\label{crowley}
\end{center}
\end{figure}

Mimicking the structure of the Levins model only in multiple dimensions, this model can also provide a simple mechanistic illustration of the impact of "environmental," e.g. exogenous, noise as well as demographic noise on a system.  We can focus on the dynamics of a single species, say, $y$, treating the role of species $x$ as some external disturbance that can make open patches temporarily uninhabitable and displace inhabited patches.  
-->

### 2018-02-27 paper/paper.Rmd
environmental noise

The van Kampen expansion can also be useful in illustrating how 
environmental noise enters a model as a separate process or dimension.
Like the expansion of demographic noise, this derivation leads to 
environmental noise that is approximately Gaussian in distribution,
but unlike the canonical model, leads to explicitly autocorrelated rather 
than white noise (see derivation, Appendix B).  This result is intuitive:
for continuous time models, all fluctations are positively auto-correlated
by definition -- measure any continuous process at ever-closer intervals in
time and these measurements must converge. As the derivation makes it clear,
it is in fact the *relative* time scales of the environmental process
and population level process that really matter.  The degree of coupling
to environmental noise also depends on the derivative of the population 
dynamics with respect to the stochastically varying environmental variable:
the larger that derivative, the stronger the contribution of environmental
fluctuations to population fluctuations. For example, if extinction
rate $e$ in our Levins' model is allowed to vary
stochastically as Gaussian white noise with variance $\sigma_e^2$
and mean $e$, we find the expected variance introduced in the
population to be:

$$\sigma_n^2 = \frac{e^2}{(c+e)^2} \sigma_e^2 + \frac{e}{c}N $$

Note that the contribution of environmental stochasticity is
independent of system size and directly proportional to the variance
in the environment.

>>>>

The van Kampen expansion can also be useful in illustrating how 
environmental noise enters a model as a separate process or dimension.
Like the expansion of demographic noise, this derivation leads to 
environmental noise that is approximately Gaussian in distribution,
but unlike the canonical model, leads to explicitly autocorrelated rather 
than white noise (see derivation, Appendix B).  This result is intuitive:
for continuous time models, all fluctations are positively auto-correlated
by definition -- measure any continuous process at ever-closer intervals in
time and these measurements must converge. As the derivation makes it clear,
it is in fact the *relative* time scales of the environmental process
and population level process that really matter.  The degree of coupling
to environmental noise also depends on the derivative of the population 
dynamics with respect to the stochastically varying environmental variable:
the larger that derivative, the stronger the contribution of environmental
fluctuations to population fluctuations. For example, if extinction
rate $e$ in our Levins' model is allowed to vary
stochastically as Gaussian white noise with variance $\sigma_e^2$
and mean $\bar e$, we find the expected variance introduced in the
population to be:

$$\sigma_n^2 = \frac{\bar e}{c}N + \frac{\left[(1-\frac{\bar{e}}{c}) N\right]^2}{(\bar e - c)^2} \sigma_e^2$$

where $\left[(1-\frac{\bar{e}}{c}) N$ is the equilibrium population size.
This reflects the assumption of the canonical equation, Eq \eqref{canonical}
that the standard deviation scales in direct proportion to the system size $N$,
while the demographic noise scales as the square root.  Consequently,
for a large system, the demographic contribution becomes proportionally smaller
while the environmental contribution remains the same.
